### 2023-02-28 calculus.tex
Updates from Overleaf
Finish marking binary products section

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not any singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
indicate a type mismatch error on the other.

@@

We now demonstrate how one might extend or adapt the marking calculus for new typing features by
adding binary product types. First, we extend the syntax of types and the accompanying notions of
consistency and join, which closely mirror the definitions for arrow types.
%
\[\begin{array}{rrcl}
  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TPair{\TMV}{\TMV}
\end{array}\]

We additionally define an auxiliary notion of \emph{matched binary product types} in
\cref{fig:calculus-matched-product}: the judgment $\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes
that $\TMV$ may be considered the binary product type $\TPair{\TMV_1}{\TMV_2}$.

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\matchedPair{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMPHole}

    \judgment{ }{
      \matchedPair{\TPair{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
    }{TMPProd}
  \end{mathpar}
  %
  \caption{Matched product type.}
  \label{fig:calculus-matched-product}
\end{figure}

The unmarked language is extended with a pair constructor and projection operators.
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots
                               \mid \EPair{\EMV}{\EMV}
                               \mid \EProjL{\EMV} \mid \EProjR{\EMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\ctx}{\EMV_1}{\TMV_1} \\
    \ctxSynType{\ctx}{\EMV_2}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TPair{\TMV_1}{\TMV_2}}
  }{USPair}

@@

  \judgment{
    \ctxSynType{\ctx}{\EMV}{\TMV} \\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\EProjR{\EMV}}{\TMV_2}
  }{USProjR} \\
\end{mathpar}
%
Similarly to application, \textsc{USProjL} and \textsc{USProjR} employ the matched binary product
type; a projection $\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be
considered to be a pair.

\input{figs/calculus-products-mexp}

\input{figs/calculus-products-marking}

>>>>

Sometimes, the type an expression \emph{should} be is known. Consider
\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
indicates a mismatch between the expected and actual types of the operand. The application in
\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
this case, however, the expectation is not a singular type, but any member of a family of function
types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
indicate a type mismatch error on the other.

@@

We now demonstrate how one might extend or adapt the marking calculus for new typing features by
adding binary product types. First, in \cref{fig:calculus-product-judgments}, we extend the syntax
of types and the accompanying notions of consistency and join, which closely mirror the definitions
for arrow types.
%
\[\begin{array}{rrcl}
  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TPair{\TMV}{\TMV}
\end{array}\]

We additionally define an auxiliary notion of \emph{matched binary product types}: the judgment
$\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
type $\TPair{\TMV_1}{\TMV_2}$.

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
  %
  \begin{mathpar}
    \judgment{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TPair{\TMV_1}{\TMV_2}}{\TPair{\TMV_1'}{\TMV_2'}}
    }{TCProd} \\
  \end{mathpar}

  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
  \[\begin{array}{rcl}
    & \cdots & \\
    \joinsTo{(\TPair{\TMV_1}{\TMV_2})}{(\TPair{\TMV_1'}{\TMV_2'})}{\TPair{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
  \end{array}\]

  \judgbox{\ensuremath{\matchedPair{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMPHole}

    \judgment{ }{
      \matchedPair{\TPair{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
    }{TMPProd}
  \end{mathpar}
  %
  \caption{Binary product consistency, join, and matched product type.}
  \label{fig:calculus-product-judgments}
\end{figure}

The unmarked language is extended with a pair constructor and projection operators. In addition to
the synthesis rules, we have explicit analysis rules. If an expected type is known for a pair
expression, the responsibility for type errors is given to the components. Likewise, we shift blame
to the projected expression when analysing projections.
%
\[\begin{array}{rrcl}
  \EMName  & \EMV  & \Coloneqq & \cdots
                               \mid \EPair{\EMV}{\EMV}
                               \mid \EProjL{\EMV} \mid \EProjR{\EMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\ctx}{\EMV_1}{\TMV_1} \\
    \ctxSynType{\ctx}{\EMV_2}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TPair{\TMV_1}{\TMV_2}}
  }{USPair}

@@

  \judgment{
    \ctxSynType{\ctx}{\EMV}{\TMV} \\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\EProjR{\EMV}}{\TMV_2}
  }{USProjR} \\

  \judgment{
    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaType{\ctx}{\EMV_1}{\TMV_1} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TMV}
  }{UAPair}

  \judgment{
    \ctxAnaType{\ctx}{\EMV}{\TPair{\TMV}{\TUnknown}}
  }{
    \ctxAnaType{\ctx}{\EProjL{\EMV}}{\TMV}
  }{UAProjL}

  \judgment{
    \ctxAnaType{\ctx}{\EMV}{\TPair{\TUnknown}{\TMV}}
  }{
    \ctxAnaType{\ctx}{\EProjR{\EMV}}{\TMV}
  }{UAProjR}
\end{mathpar}
%
Similarly to application, \textsc{USProjL} and \textsc{USProjR} employ the matched binary product
type judgment; a projection $\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may
be considered to be a pair.

The marked language is extended with the same forms, and we devise the following marking rules,
again modelling off the above unmarked typing rules:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots
                               \mid \ECPair{\ECMV}{\ECMV}
                               \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                               \mid \ECInconMatchedPair{\ECMV}
\end{array}\]
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
  }{ISPair}

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
  }{ISProjL1}

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
  }{ISProjR1} \\

  \judgment{
    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{IAPair1}

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TPair{\TMV}{\TUnknown}}
  }{
    \ctxAnaFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV}
  }{IAProjL}

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TPair{\TMV}{\TUnknown}}
  }{
    \ctxAnaFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV}
  }{IAProjR}
\end{mathpar}
%
accompanied by corresponding typing rules for marked expressions:
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\ctx}{\ECMV_1}{\TMV_1} \\
    \ctxSynType{\ctx}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
  }{MSPair}

  \judgment{
    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
  }{MSProjL1}

  \judgment{
    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECProjR{\ECMV}}{\TMV_2}
  }{MSProjR1}

  \judgment{
    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaType{\ctx}{\ECMV_1}{\TMV_1} \\
    \ctxAnaType{\ctx}{\ECMV_2}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
  }{MAPair1}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV}{\TPair{\TMV}{\TUnknown}}
  }{
    \ctxAnaType{\ctx}{\ECProjL{\ECMV}}{\TMV}
  }{MAProjL}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV}{\TPair{\TUnknown}{\TMV}}
  }{
    \ctxAnaType{\ctx}{\ECProjR{\ECMV}}{\TMV}
  }{MAProjR}
\end{mathpar}

However, as with lambda abstractions and application seen previously, two sets of rules are further
necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
binary product may fail, requiring marking and typing rules to handle that case:
%
\begin{mathpar}
  \judgment{
    \notMatchedPair{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECInconType{\ECPair{\ECMV_1}{\ECMV_2}}}{\TMV}
  }{IAPair2}

  \judgment{
    \notMatchedPair{\TMV} \\
    \ctxAnaType{\ctx}{\ECMV_1}{\TUnknown} \\
    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECPair{\ECMV_1}{\ECMV_2}}}{\TMV}
  }{MAPair2}
\end{mathpar}

The other kind of error may arise when the subject of a projection does not synthesize a matched
product type. As in our development of marking for applications, we may place the subject in an
inconsistent type mark, or we may treat it as a special ``constrained synthesis'' mode, utilising a
new error mark $\ECInconMatchedPair{\ECMV}$. The latter approach leads to the following rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
    \notMatchedPair{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECInconMatchedPair{\ECMV}}}{\TMV_1}
  }{ISProjL2}

  \judgment{
    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
    \notMatchedPair{\TMV}
  }{
    \ctxSynType{\ctx}{\ECProjL{\ECInconMatchedPair{\ECMV}}}{\TUnknown}
  }{MSProjL2} \\

  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
    \notMatchedPair{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECInconMatchedPair{\ECMV}}}{\TMV_2}
  }{ISProjR2}

  \judgment{
    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
    \notMatchedPair{\TMV}
  }{
    \ctxSynType{\ctx}{\ECProjR{\ECInconMatchedPair{\ECMV}}}{\TUnknown}
  }{MSProjR2}
\end{mathpar}

\input{let}
